---
title: "A Site In Progress"
---

\
\
\
Hello, my name is Shahria but most people call me Shah. Welcome to my site! Here's a little bit about me.\

I'm a data scientist on Wall Street, working on AI development and testing in equity research. Currently I'm getting my Master's at CUNY Baruch. But in between work and school, I take a lot of pictures. Sunsets, cityscapes, portraits, nature, astrophotography — I love it all! My interest started back in school when I worked on an in-depth research report on Alzheimer's. It's a brutal and unforgiving disease with no cure. I wanted to remember all the things I've experienced when I became old, so I started taking pictures with my iPhone 6 at the time.

My current setup:

* Fujifilm X100V for travel photography, compactness and film recipes
* Sony A7RIII + Zeiss 85mm lens for portraits and everything else \


Thanks for visiting!



```{r}
#| include: false
1+1
```
--------------
Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`


```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude  <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "This is where I go to school ! <b>Baruch College</b>!")
```