---
title: "Labor Statistics"
format: 
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-location: right
    toc-depth: 3
    smooth-scroll: true
    anchor-sections: true
    css: mp02.css
---

Hello! Today we will be looking at data from the Bureau of Labor Statistics. This data is released on the first Friday of every month and is highly anticipated by a variety of players in the stock market. The BLS produces detailed industry estimates of nonfarm employment, hours, and earnings of workers on payrolls. Ultimately this estimate of jobs gives us a gauge of how well the economy is doing which can have second order effects on stocks and major indices.

Click on the code chunks below to expand or collapse code if you're following along. Use the navigation bar on the right to navigate through the analysis and policy brief. 

For this analysis, we will go to the BLS website and get monthly data going as far back as 1979. This requires a POST request instead of a simple GET. The following code block visits the website, scrapes the data and provides a well formatted output in the form of a datatable. Click on the code block to view the syntax.

# BLS Seasonally Adjusted Employment Numbers
```{r}
#| code-fold: true
#| code-summary: "historical jobs datapull"
#| message: false
#| warning: false

# load all required libraries
library(httr2)
library(readr)
library(rvest)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(DT)

### 1. define and perform the http request
dir_path <- "data/mp04"
csv_path <- file.path(dir_path, "ces_historical_data.csv")

# create directory if it doesn't exist
if (!dir.exists(dir_path)) {
  dir.create(dir_path, recursive = TRUE)
}

# if cached file exists, read and load it
if (file.exists(csv_path)) {
  ces_data <- read_csv(csv_path, show_col_types = FALSE)  
} else {
  # otherwise download the data
  message("downloading ces data from bls...")
  
  bls_url <- "https://data.bls.gov/pdq/SurveyOutputServlet"
  form_payload <- list(
    "series_id" = "CES0000000001",
    "from_year" = "1979",
    "to_year" = "2025",
    "survey" = "CE"
  )
  
  # build and perform the request
  resp <- request(bls_url) |>
    req_body_form(!!!form_payload) |>
    req_perform() |> 
    resp_check_status() 
  
  ###  parse the html response with rvest
  html_content <- resp_body_html(resp)
  
  # extract the main data table
  raw_table <- html_content |>
    html_element("table#table0") |>
    html_table()
  
  ###  pivot and clean the data
  ces_data <- raw_table |>
    filter(!is.na(Year) & !str_detect(Year, "Preliminary")) |>
    
    pivot_longer(
      cols = -Year,
      names_to = "month",
      values_to = "level_str"
    ) |>
    
    #  modify column types and finalize for table
    mutate(date_str = str_c(Year, " ", month)) |>
    mutate(date = suppressWarnings(ym(date_str))) |>
    mutate(level = as.numeric(str_replace_all(level_str, ",", ""))) |>
    
    drop_na(date, level) |>
    
    select(date, level) |>
    arrange(date)
  
  # save to file
  write_csv(ces_data, csv_path)
  message("ces data saved to ", csv_path)
}
```

Let's see how this looks in a clean datatable
```{r}
#| code-fold: true
#| code-summary: "show datatable"
#| message: false
#| warning: false

### display as an interactive datatable
ces_data |>
  mutate(
    Date = format(date, "%Y-%m"),  # format date 
    Employees = format(level, big.mark = ",")  # add commas to clean up numbers
  ) |>
  select(Date, Employees) |>
  datatable(
    rownames = FALSE,
    options = list(
      pageLength = 10, #10 results 
      lengthMenu = c(10, 25, 50, 100),
      order = list(list(0, 'desc')),  # sort by date descending
      dom = 'lfrtip'
    ),
    caption = "CES Total Nonfarm Payroll Employment (Seasonally Adjusted, thousands)", #title
    filter = 'top',
    class = 'display'
  ) |>
  formatStyle(
    'Employees',
    fontWeight = 'bold'
  )
```

# BLS Seasonally Adjusted Revisions

This data gets revised after the initial release as more accurate metrics come in. Let's take a look at the revisions for each month, going back to 1979. The following code blocks pull the revisions data and format it in a clean datatable.
```{r}
#| code-fold: true
#| code-summary: "revisions datapull"
#| message: false
#| warning: false

library(httr2)
library(rvest)
library(readr)
library(dplyr)
library(lubridate)
library(purrr)
library(DT)
library(readr)

# function to get ces revisions with saved file
get_ces_revisions <- function() {
  dir_path <- "data/mp04"
  csv_path <- file.path(dir_path, "ces_revisions_data.csv")
  
  # create directory if it doesn't exist
  if (!dir.exists(dir_path)) {
    dir.create(dir_path, recursive = TRUE)
  }
  
  # if saved file exists, read and return it
  # # otherwise, download the data
  if (file.exists(csv_path)) {
    return(read_csv(csv_path, show_col_types = FALSE))
  }

  # helper function to extract revision data for a single year
  extract_year_revisions <- function(year, html_content) {
    # table ids are just the year 
    table_id <- paste0("#", year)
    
    # extract the table body only 
    table_data <- html_content |>
      html_element(table_id) |>
      html_element("tbody") |>
      html_table(header = FALSE)
    
    # take first 12 rows which are months and select relevant columns
    # column 1 is month, column 3 is 1st estimate, column 5 is 3rd estimate
    table_data |>
      slice(1:12) |>
      select(month = 1, original = 3, final = 5) |>
      mutate(
        # create date from year and month
        date = ym(paste(year, month)),
        # clean and convert to numeric value
        original = as.numeric(gsub(",", "", original)),
        final = as.numeric(gsub(",", "", final)),
        # calculate revision
        revision = final - original
      ) |>
      select(date, original, final, revision) |>
      drop_na()  # remove any rows that didn't read correctly
  }
  
  # make the request following the same pattern as bls function. copy function from mp02
  response <- request("https://www.bls.gov") |>
    req_url_path("web", "empsit", "cesnaicsrev.htm") |>
    req_headers(
      `User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:143.0) Gecko/20100101 Firefox/143.0"
    ) |>
    req_error(is_error = \(resp) FALSE) |>
    req_perform()
  
  # check status
  resp_check_status(response)
  html_content <- resp_body_html(response)
  
  # apply function to all years from 1979 to 2025 and combine
  ces_revisions <- map(
    1979:2025, 
    .progress = TRUE,
    possibly(~extract_year_revisions(.x, html_content), otherwise = NULL) #accounts for errors
  ) |>
    list_rbind()
  
  # save to directory
  write_csv(ces_revisions, csv_path)
  return(ces_revisions)
}

# load the data and call function
ces_revisions <- get_ces_revisions()
```

```{r}
#| code-fold: true
#| code-summary: "show datatable"
#| message: false
#| warning: false
# apply function to all years from 1979 to 2025 and combine

# create a nice datatable
ces_revisions |>
  mutate(
    Date = format(date, "%Y-%m"), #format date
    Original = format(original, big.mark = ","), #account for commas
    Final = format(final, big.mark = ","),
    Revision = format(revision, big.mark = ",")
  ) |>
  select(Date, Original, Final, Revision) |>
  datatable(
    rownames = FALSE,
    options = list(
      pageLength = 10, #10 rows
      lengthMenu = c(10, 25, 50, 100),
      order = list(list(0, 'desc')),
      dom = 'lfrtip'
    ),
    caption = "CES Total Nonfarm Employment Revisions (thousands)", #title
    filter = 'top',
    class = 'display'
  )
```

# Exploratory Analysis

From here, we will combine these two tables from BLS and perform analysis. The code chunk below creates a join
```{r}
#| code-fold: true
#| code-summary: "show table join"
#| message: false
#| warning: false

#load packages
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(scales)
library(DT)

# join the two tables together
ces_combined <- ces_data |>
  inner_join(ces_revisions, by = "date") |>  # simplified join syntax
  mutate(
    year = year(date), #convert to date
    month_name = month(date, label = TRUE),
    decade = floor(year / 10) * 10,
    # calculate relative metrics
    revision_pct = (revision / final) * 100,
    abs_revision = abs(revision), #absolute
    abs_revision_pct = abs(abs_revision / final) * 100, #absolute percentage
    revision_as_pct_employment = (abs_revision / level) * 100, #$revision
    is_positive_revision = revision > 0
  )

# display in a datatable
ces_combined |>
  select(date, level, original, final, revision, revision_pct, abs_revision_pct) |>  # select specific columns
  mutate(
    Date = format(date, "%Y-%m"), #format date
    `Employment Level` = format(level, big.mark = ","), #format with commas
    `Original Estimate` = format(original, big.mark = ","),
    `Final Estimate` = format(final, big.mark = ","),
    `Revision` = format(revision, big.mark = ","),
    `Revision %` = round(revision_pct, 2),
    `Absolute Revision %` = round(abs_revision_pct, 2)
  ) |>
  select(Date, `Employment Level`, `Original Estimate`, `Final Estimate`,  #keep these columns for table
         `Revision`, `Revision %`, `Absolute Revision %`) |>
  datatable(
    rownames = FALSE,
    options = list(
      pageLength = 10, #10 rows
      lengthMenu = c(10, 25, 50, 100),
      order = list(list(0, 'desc')),
      dom = 'lfrtip'
    ),
    caption = "CES Total Nonfarm Employment and Revisions (thousands)",  #title
    filter = 'top',
    class = 'display'
  )

```

### Statistics

Let's look at the mean and median revisions over the last 45 years. These come out to **11,500** and **10,000**, respectively. That's not bad, compared to some of the massive revisions. For a country of 300 million people, the estimates are pretty accurate.

A positive mean revision means the BLS generally underestimated job growth in the initial report, requiring a subsequent upward correction. A negative mean revision means the BLS generally overestimated job growth, requiring a downward correction.
```{r}
#| code-fold: true
#| code-summary: "show summary stats"
#| message: false
#| warning: false

#load package
library(DT)
library(dplyr)

# overall revision statistics
overall_stats <- ces_combined |>
  summarise(
    mean_revision = mean(revision, na.rm = TRUE), #mean
    median_revision = median(revision, na.rm = TRUE) #mean 
  )

overall_stats |>
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") |> #pivot table
  mutate(
    Metric = case_when( #rename columns
      Metric == "mean_revision" ~ "Mean Revision",
      Metric == "median_revision" ~ "Median Revision",
    ),
    Value = round(Value, 2)
  ) |>
  datatable(
    rownames = FALSE,
    options = list(dom = 't', ordering = FALSE),
    caption = "Overall Revision Statistics" #title
  )
```


Let's look at the anomalies in revision, sorted by date. Each of these revisions correspond to important economic events. In the early 80s, there was high inflation and subsequently high rates by Volcker to combat said inflation. The largest revision in modern day is due to Covid and its Delta/Omicron variants.
```{r}
#| code-fold: true
#| code-summary: "show largest revision"
#| message: false
#| warning: false

# largest revisions in ces history
largest_revisions <- ces_combined |>
  arrange(desc(abs_revision)) |> #sort by absolute revision
  select(date, original, final, revision, level) |>
  head(10) #return 10 rows

largest_revisions |>
  arrange(date) |>  # sort by date first, before mutating
  mutate(
    Date = format(date, "%Y-%m"), #format date
    `Original Estimate` = format(original, big.mark = ","), #format with commas
    `Final Estimate` = format(final, big.mark = ","),
    `Revision` = format(revision, big.mark = ","),
    `Employment Level` = format(level, big.mark = ",")
  ) |>
  select(Date, `Original Estimate`, `Final Estimate`, `Revision`, `Employment Level`) |> #select these columns
  datatable(
    rownames = FALSE,
    options = list(dom = 't', ordering = FALSE),
    caption = "Top 10 Largest Absolute Revisions" #title
  )
```


Now let's look at the most positive and negative revisions in history. Most positive revisions come after a month of unexpected data and signal a recovery in the economy. Whereas most negative revisions are actually lagging and come after a major shock in the economy. We didn't know the revised numbers for the Covid shock until a few months later when the magnitude of the pandemic was starting to be realized.
```{r}
#| code-fold: true
#| code-summary: "show postive/negative revisions"
#| message: false
#| warning: false

#. most positive and most negative revisions
extreme_revisions <- ces_combined |>
  arrange(desc(revision)) |> #sort by revision
  select(date, original, final, revision) |>
  slice(c(1:5, (n()-4):n())) #select top 5 and bottom 5

extreme_revisions |>
  mutate(
    Date = format(date, "%Y-%m"), #format date and numbers by commas
    `Original Estimate` = format(original, big.mark = ","),
    `Final Estimate` = format(final, big.mark = ","),
    `Revision` = format(revision, big.mark = ","),
    Type = ifelse(row_number() <= 5, "Most Positive", "Most Negative")
  ) |>
  select(Type, Date, `Original Estimate`, `Final Estimate`, `Revision`) |> #select these columns
  datatable(
    rownames = FALSE,
    options = list(dom = 't', ordering = FALSE),
    caption = "Most Extreme Revisions (Top 5 Positive and Negative)"
  ) |> #title
  formatStyle('Type',
    backgroundColor = styleEqual(c('Most Positive', 'Most Negative'), c('#d4edda', '#f8d7da')) #color coordinate highest and lowest revisions
  )
```


How do the revisions fare by decade? We start to notice a pattern in the decades using average revision. We oscillate between overestimation and underestimation of jobs, starting from the 70s all the way to modern day. It seems like for every decade that we under or overshoot, we adjust the following decade and the estimates become more refined. While the mean absolute revision **(86.9k)** is very high, driven primarily by the extreme volatility of the 2020 pandemic lockdowns and recovery, the fact that the mean is near zero implies that the BLS has become more accurate at employment estimation
```{r}
#| code-fold: true
#| code-summary: "show revisions by decade"
#| message: false
#| warning: false

# revisions by decade
decade_stats <- ces_combined |>
  group_by(decade) |>
  summarise( #summarize median and mean stats
    n_months = n(),
    mean_revision = mean(revision, na.rm = TRUE),
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    mean_abs_revision_pct = mean(abs_revision_pct, na.rm = TRUE),
    pct_positive = mean(is_positive_revision, na.rm = TRUE) * 100,
    .groups = "drop"
  )
#show results in nice datatable
decade_stats |>
  mutate( 
    Decade = paste0(decade, "s"),
    `# Months` = n_months, 
    `Mean Revision` = round(mean_revision, 1), #mean
    `Mean Abs Revision` = round(mean_abs_revision, 1), #mean abs revision
    `Mean Abs Revision (%)` = round(mean_abs_revision_pct, 2),
    `% Positive` = round(pct_positive, 1)
  ) |>
  select(Decade, `# Months`, `Mean Revision`, `Mean Abs Revision`, 
         `Mean Abs Revision (%)`, `% Positive`) |> #select these columns
  datatable(
    rownames = FALSE,
    options = list(dom = 't', ordering = FALSE),
    caption = "Revision Statistics by Decade"
  )
```

Can we identify any seasonality in the data if we break it down by month? We know businesses have hiring cycles, so it should be reflected in the data. But are people more likely to be laid off within a specific month? Based on our data, it seems that hiring slows down in the Summertime and Q1. Revisions increase during Q2 and Q4
```{r}
#| code-fold: true
#| code-summary: "show seasonality stats"
#| message: false
#| warning: false

#  revisions by month of year
month_stats <- ces_combined |>
  group_by(month_name) |>
  summarise( #summarize stats like mean revisions
    n_observations = n(),
    mean_revision = mean(revision, na.rm = TRUE), #mean revision
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    mean_abs_revision_pct = mean(abs_revision_pct, na.rm = TRUE),
    .groups = "drop"
  )

month_stats |>
  mutate( #show values in a nice formatted datatable
    Month = as.character(month_name),
    `# Observations` = n_observations, #mean revisions
    `Mean Revision` = round(mean_revision, 1),
    `Mean Abs Revision` = round(mean_abs_revision, 1),
    `Mean Abs Revision (%)` = round(mean_abs_revision_pct, 2)
  ) |>
  select(Month, `# Observations`, `Mean Revision`, 
         `Mean Abs Revision`, `Mean Abs Revision (%)`) |> #select these columns
  datatable(
    rownames = FALSE,
    options = list(dom = 't', ordering = FALSE),
    caption = "Revision Statistics by Month" #title
  )
```


How have revisions changed over time? Excluding 2021 due to COVID, we can notice that the _magnitude_ of revisions has decreased in recent years compared to historical data. This signals that the models and estimates used by BLS has gotten better at predicting employment.
```{r}
#| code-fold: true
#| code-summary: "show revision over time"
#| message: false
#| warning: false

# 6. how has revision accuracy changed over time
trend_stats <- ces_combined |>
  group_by(year) |>
  summarise( #get mean revision and percentage by year
    mean_abs_revision = mean(abs_revision, na.rm = TRUE), 
    mean_abs_revision_pct = mean(abs_revision_pct, na.rm = TRUE),
    pct_positive = mean(is_positive_revision, na.rm = TRUE) * 100,
    .groups = "drop"
  )

#show data in nice formatted table
bind_rows(head(trend_stats, 5), tail(trend_stats, 5)) |>
  mutate(
    Year = year,
    `Mean Abs Revision` = round(mean_abs_revision, 1), #rund mean to 1 decimal place
    `Mean Abs Revision (%)` = round(mean_abs_revision_pct, 2), #round percent to 2 decimal places
    `% Positive` = round(pct_positive, 1),
    Period = ifelse(row_number() <= 5, "Early Years (1979-1983)", "Recent Years (2021-2025)")
  ) |>
  select(Period, Year, `Mean Abs Revision`, `Mean Abs Revision (%)`, `% Positive`) |> #select these columns
  datatable(
    rownames = FALSE,
    options = list(dom = 't', ordering = FALSE),
    caption = "Revision Trends: First and Last 5 Years"
  ) |>
  formatStyle('Period',
    backgroundColor = styleEqual(c('Early Years (1979-1983)', 'Recent Years (2021-2025)'), 
c('#fff3cd', '#d1ecf1')) #color coordinate by years
  )
```


### Visuals

Now let's look at some visuals

We can distinctly notice a shift post Covid in revisions. There are a lot more downward revisions than upwards, signaling that perhaps employment has fundamentally changed. It could be a number of things: there's more layoffs due to economic weakening, AI related increases in productivity, difficulty for younger workers to break into their respective fields.
```{r}
#| code-fold: true
#| code-summary: "show revisions code"
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Timeline of CES Revisions Over 45 Years"

# timeline of revisions with color coding
ggplot(ces_combined, aes(x = date, y = revision, fill = is_positive_revision)) +
  geom_col(alpha = 0.7) + #adjust the transparency of plot
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_fill_manual(
    values = c("TRUE" = "darkgreen", "FALSE" = "darkred"), #green for positive revisions, red for negative revisions
    labels = c("Downward Revision", "Upward Revision"),
    name = ""
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") + #break at 5 years for scale
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  labs( #labels such as title, subtitle, x and y axis
    title = "CES Monthly Revisions Over Time",
    subtitle = "Difference between initial and final employment estimates (thousands)",
    x = "Date",
    y = "Revision (thousands)",
    caption = "Positive values indicate upward revisions; Negative indicate downward revisions"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```


The next graph shows how the magnitude of revisions has decreased over time by decade. We notice that the outliers in 2020 resemble the 1980s, which checks out. Both decades have had a lot of external shocks and you'll often notice 20s inflation being compared to the 80s in news headlines. 
```{r}
#| code-fold: true
#| code-summary: "show decade comparison code"
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Revision Patterns by Decade"

# boxplot of revisions by decade
ggplot(ces_combined, aes(x = factor(decade), y = revision)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) + #adjust transparency
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  labs( #adjust title, subtitle, xand y axis
    title = "Distribution of CES Revisions by Decade",
    subtitle = "Box plots show median, quartiles, and outliers",
    x = "Decade",
    y = "Revision (thousands)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )
```

The following graph shows revisions by month. We notice that September typically has the largest revisions, followed by April and March, respectively. Coincidentally, September is the worst performing month in the stock market. It even has its own name: The September Effect. During this time, major indices usually underperform. We can't exactly assume it's due to employment but it may play a part.
```{r}
#| code-fold: true
#| code-summary: "show monthly pattern code"
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Systematic Monthly Patterns in Revision Magnitude"

# are certain months systematically revised more
ggplot(month_stats, aes(x = month_name, y = mean_abs_revision)) +
  geom_col(fill = "coral", alpha = 0.8) + #adjust transparency 
  geom_text(aes(label = round(mean_abs_revision, 1)),  #round value to 1 decimal place
            vjust = -0.5, size = 3) +
  theme_minimal() +
  labs( #title, subtitle, x and y axis values
    title = "Average Absolute Revision by Month",
    subtitle = "Do certain months have systematically larger revisions?",
    x = "Month",
    y = "Mean absolute revision (thousands)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

We noticed this earlier in the first graph, but this helps illustrate it better. The number of positive revisions has plummeted post 2020. This isn't good. The last five years may feel like constant economic stress for a lot of people. Well that's probably because it is. Employment hasn't fully recovered and people are struggling. In contrast, we notice a big bounce in positive revisions during and after the housing crisis from 2005 to 2010. We bounced back far better back then compared to now. 
```{r}
#| code-fold: true
#| code-summary: "show positive/negative trend code"
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Proportion of Positive vs Negative Revisions Over Time"

# proportion of positive revisions over time
ces_combined |>
  mutate(year_group = floor(year / 5) * 5) |>
  group_by(year_group) |> 
  summarise( #group by year and get average revision 
    pct_positive = mean(is_positive_revision, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = year_group, y = pct_positive)) + #x and y axis values
  geom_line(color = "darkblue", linewidth = 1) + #dark blue for box
  geom_point(color = "darkblue", size = 3) + 
  geom_hline(yintercept = 50, linetype = "dashed", color = "red", alpha = 0.5) + #adjust transparency 
  scale_x_continuous(breaks = seq(1980, 2025, 5)) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 10)) +
  theme_minimal() +
  labs( #title, subtitle x and y axis
    title = "Percentage of Positive CES Revisions Over Time",
    subtitle = "5-year rolling averages; dashed line shows 50% (no systematic bias)",
    x = "5-year Period",
    y = "Percentage of Revisions That Were Positive (%)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )
```


The graph below shows employment over time. It has steadily increased, despite shocks to the workforce. One thing we notice is that employment has recovered post COVID, but the magnitude of revisions is approaching 2008 levels. So unfortunately, it's difficult to accurately conclude the recovery of employment post pandemic.
```{r}
#| code-fold: true
#| code-summary: "show employment/revisions dual axis"
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "CES Employment Levels and Absolute Revisions Over Time"

# create scaling factor based on data ranges
coeff <- max(ces_combined$level) / max(ces_combined$abs_revision)

ggplot(ces_combined, aes(x = date)) +
  # employment line (primary axis)
  geom_line(aes(y = level), 
            color = "steelblue", #blue for employment
            linewidth = 1.2,
            alpha = 0.8) + #adjust transparency 
  
  # absolute revision line (scaled to match employment range)
  geom_line(aes(y = abs_revision * coeff), 
            color = "coral", #coral for revision
            linewidth = 1,
            alpha = 0.8) + #adjust transparency 
  
  # dual y-axis
  scale_y_continuous(
    name = "Employment (thousands)",
    labels = comma,
    sec.axis = sec_axis(
      ~ . / coeff,
      name = "Absolute Revision (thousands)",
      labels = comma
    )
  ) +
  
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  
  theme_minimal() +
  
  labs( #title, subtitle, x axis and caption
    title = "CES Employment Level and Absolute Revisions Over Time",
    subtitle = "Blue Line: Total Employment (left axis) | Coral Line: Absolute Revision Magnitude (right axis)",
    x = "Date",
    caption = "Higher coral line indicates larger revisions between initial and final estimates"
  ) +
  
  theme( #adjust color and font
    plot.title = element_text(face = "bold", size = 14),
    axis.title.y = element_text(color = "steelblue", face = "bold"),
    axis.title.y.right = element_text(color = "coral", face = "bold"),
    axis.text.y = element_text(color = "steelblue"),
    axis.text.y.right = element_text(color = "coral"),
    legend.position = "none"
  )
```

Let's zoom in on the last two years. We can approximate the actual employment numbers by taking the total employment numbers and subtracting the revisions from it. Plotting this out, we notice that the gap between projected employment and actual employment is widening. As a matter of fact, the most recent gap is over 100,000 jobs. That's very concerning!
```{r}
#| code-fold: true
#| code-summary: "show 2024-2025 original vs revised employment"
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Original vs Revised Employment Estimates (2024-2025)"

# filter data for 2024-2025 and calculate revised employment
recent_data <- ces_combined |>
  filter(year >= 2024) |>
  mutate(
    # original estimate is final minus revision
    original_employment = final,
    # the level is the final estimate
    revised_employment = level
  )

# create the plot with both employment lines
ggplot(recent_data, aes(x = date)) +
  # original estimate line
  geom_line(aes(y = original, color = "Original Estimate"), 
            linewidth = 1.2, 
            linetype = "dashed",
            alpha = 0.8) + #ADJUST transparency 
  geom_point(aes(y = original, color = "Original Estimate"), 
             size = 3, alpha = 0.8) +
  
  # final estimate line
  geom_line(aes(y = final, color = "Final Estimate"), 
            linewidth = 1.2) +
  geom_point(aes(y = final, color = "Final Estimate"), 
             size = 3) +
  
  # add ribbons to show the revision gap
  geom_ribbon(aes(ymin = pmin(original, final), #set y axis min and max values
                  ymax = pmax(original, final)),
              fill = "gray", alpha = 0.3) +
  
  # color scheme
  scale_color_manual(
    name = "",
    values = c("Original Estimate" = "coral",  #set color scheme for two plots
               "Final Estimate" = "steelblue")
  ) +
  
  scale_y_continuous( #y value label
    name = "Employment (thousands)",
    labels = comma
  ) +
  
  scale_x_date(
    date_breaks = "1 month", 
    date_labels = "%b\n%Y",
    expand = c(0.02, 0)
  ) +
  
  theme_minimal() +
  
  labs( #title subtitle x axis and caption
    title = "CES Employment: Original vs Final Estimates (2024-2025)",
    subtitle = "Gray shaded area shows the magnitude of revisions between first and third estimates",
    x = "",
    caption = "Coral dashed line: First published estimate | Blue solid line: Final revised estimate"
  ) +
  
  theme(#adjust font and text
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

# Statical Analysis 

Earlier we noticed that the number of negative revisions have increased post 2020. From here on, we'll perform statistical tests to confirm that this is in fact true and not just simply randomness. 

```{r}
#| code-fold: true
#| code-summary: "show statistical inference tests"
#| message: false
#| warning: false

#install infer package if not installed
if (!require("infer", quietly = TRUE)) {
  # set CRAN mirror before installing
  options(repos = c(CRAN = "https://cloud.r-project.org/"))
  install.packages("infer")
  library(infer)
}

library(dplyr)
library(DT)

# prepare data for tests
ces_test_data <- ces_combined |>
  mutate(
    # create binary variables for tests
    post_2000 = year >= 2000, #after 2000
    post_2020 = year >= 2020, #after 2020
    negative_revision = revision < 0,
    large_revision = abs_revision_pct > 1,
    # categorize employment changes
    employment_change = level - lag(level),
    large_employment_change = abs(employment_change) > 200
  )
```

```{r}
#| code-fold: true
#| code-summary: "show statistical inference tests"
#| message: false
#| warning: false

# test if negative revisions are more common after 2000
test1_result <- ces_test_data |>
  prop_test(negative_revision ~ post_2000, 
            order = c("TRUE", "FALSE"),
            alternative = "two.sided")
#combine results in a nice datatable. 
#use chi square and p value test
#if p value less than 0.05, then it is statistically significant
test1_result |>
  mutate(
    Test = "Negative Revisions: Post-2000 vs Pre-2000",
    `Chi-Square Statistic` = round(statistic, 3), 
    `P-Value` = format.pval(p_value, digits = 3),
    `Difference (%)` = round((upper_ci - lower_ci) / 2 * 100, 2),
    Conclusion = ifelse(p_value < 0.05,  #threshold
                       "Statistically significant difference", 
                       "No significant difference")
  ) |>
  select(Test, `Chi-Square Statistic`, `P-Value`, `Difference (%)`, Conclusion) |> #select these columns
  datatable(
    rownames = FALSE,
    options = list(dom = 't'),
    caption = "Test 1: Proportion Test for Negative Revisions"
  )
```

```{r}
#| code-fold: true
#| code-summary: "show statistical inference tests"
#| message: false
#| warning: false

# test if large revisions are more common after 2020
test2_result <- ces_test_data |>
  filter(!is.na(large_revision)) |>
  prop_test(large_revision ~ post_2020,
            order = c("TRUE", "FALSE"),
            alternative = "two.sided")

#combine results in a nice datatable. 
#use chi square and p value test
#if p value less than 0.05, then it is statistically significant
test2_result |>
  mutate(
    Test = "Large Revisions (>1%): Post-2020 vs Pre-2020",
    `Chi-Square Statistic` = round(statistic, 3),
    `P-Value` = format.pval(p_value, digits = 3),
    `Difference (%)` = round((upper_ci - lower_ci) / 2 * 100, 2),
    Conclusion = ifelse(p_value < 0.05,  #threshold
                       "Statistically significant difference", 
                       "No significant difference")
  ) |>
  select(Test, `Chi-Square Statistic`, `P-Value`, `Difference (%)`, Conclusion) |> #select these values
  datatable(
    rownames = FALSE,
    options = list(dom = 't'),
    caption = "Test 2: Proportion Test for Large Revisions"
  )
```

Well, it turns out that it is in fact random. Both large revisions and negative revisions after 2020 are not statistically significant. Despite visually appearing to show a pattern, there is no underlying pattern in the data. This is a good sanity check before crafting a narrative that supports our bias.

However, if we do a two sample t-test comparing absolute revisions pre and post 2020, we do notice statistical significance. Post 2020, the revisions are larger on an absolute basis. 
```{r}
#| code-fold: true
#| code-summary: "show statistical inference tests"
#| message: false
#| warning: false

# two-sample t-test comparing absolute revisions before and after 2020
test4_result <- ces_test_data |>
  mutate(period = ifelse(post_2020, "Post-2020", "Pre-2020")) |>
  t_test(abs_revision ~ period,
         order = c("Post-2020", "Pre-2020"),
         alternative = "two.sided")

test4_result |>
  mutate(
    Test = "Absolute Revision: Post-2020 vs Pre-2020",
    `T-Statistic` = round(statistic, 3),
    `Degrees of Freedom` = round(t_df, 0),
    `P-Value` = format.pval(p_value, digits = 3),
    `Mean Difference` = round(estimate, 2),
    `95% CI Lower` = round(lower_ci, 2),
    `95% CI Upper` = round(upper_ci, 2),
    Conclusion = ifelse(p_value < 0.05,
                       ifelse(estimate > 0, 
                              "Post-2020 revisions are significantly larger",
                              "Pre-2020 revisions were significantly larger"),
                       "No significant difference between periods")
  ) |>
  select(Test, `T-Statistic`, `Degrees of Freedom`, `P-Value`, 
         `Mean Difference`, `95% CI Lower`, `95% CI Upper`, Conclusion) |>
  datatable(
    rownames = FALSE,
    options = list(dom = 't', scrollX = TRUE),
    caption = "Test 4: Two-Sample T-Test for Absolute Revisions"
  )
```

# Fact Checking Politicians

"A lie can travel halfway around the world while the truth is still putting on its shoes" -Unknown

Modern politics in the United States can be headache inducing. Anyone can say anything. So who do you believe? It's easy to take data, twist it to fit your narrative and run wild with it. We are living in a time where fact checking is crucial. When it comes to jobs data and general data coming from the government, it is imperative that it is accurate and unbiased. Lot's of decisions are made based on this data. Especially decisions involving money. So let's take a look at some statements made by politicians using BLS jobs data and fact check.

```{r}
#| label: setup-factcheck
#| code-fold: true
#| code-summary: "setup for fact checks"
#| message: false
#| warning: false

library(tidyverse)
library(infer)
library(DT)
library(scales)

# create presidents/party data
presidents_party <- tidyr::expand_grid(year=1979:2025, 
                                       month = month.name, 
                                       president = NA, 
                                       party = NA) |> 
    mutate(president = case_when(
        (month == "January")  & (year == 1979) ~ "Carter",
        (month == "February") & (year == 1981) ~ "Reagan",
        (month == "February") & (year == 1989) ~ "Bush 41",
        (month == "February") & (year == 1993) ~ "Clinton",
        (month == "February") & (year == 2001) ~ "Bush 43",
        (month == "February") & (year == 2009) ~ "Obama",
        (month == "February") & (year == 2017) ~ "Trump I",
        (month == "February") & (year == 2021) ~ "Biden",
        (month == "February") & (year == 2025) ~ "Trump II",
    )) |>
    tidyr::fill(president) |>
    mutate(party = if_else(president %in% c("Carter", "Clinton", "Obama", "Biden"), 
                           "D", 
                           "R"))

# join with our ces data
ces_with_president <- ces_combined |>
  mutate(month = month.name[month(date)]) |>
  left_join(presidents_party, by = c("year", "month"))
```

#### Fact Check #1: Trump's 160 Million Jobs Claim
"Very close to 160 million people are now working in the United States. Nobody's ever even come close to that number." - Donald Trump, January 14, 2020


This is pretty straightforward and easy to check. 

Employment in January 2020 was 152,031 million, not "very close to 160 million". That's off by nearly 8 million jobs. However, employment did reach 160+ million in February and March 2020 before COVID

The claim that "nobody's ever even come close" is misleading, given that population has steadily increased over the years so naturally employment will go up proportionately. While the absolute number was a record, it was a natural continuation of post-recession growth trends

The 152 million number corroborates with PolitiFact and we can therefore conclude that this is mostly true. 
```{r}
#| label: factcheck1-stats
#| code-fold: true
#| code-summary: "show analysis of 160M jobs claim"

# get employment in January 2020
jan_2020_employment <- ces_data |>
  filter(date == "2020-01-01") |>
  pull(level)

# find all months that hit 160M before January 2020
months_over_160m <- ces_data |>
  filter(date <= "2020-01-01", level >= 160000) |>
  arrange(date)

# show the months
months_over_160m |>
  mutate(
    Date = format(date, "%B %Y"),
    `Employment (thousands)` = format(level, big.mark = ","),
    `Employment (millions)` = round(level / 1000, 1)
  ) |>
  select(Date, `Employment (thousands)`, `Employment (millions)`) |>
  datatable(
    rownames = FALSE,
    options = list(pageLength = 10, dom = 't'),
    caption = "All Months with Employment ≥ 160 Million Before January 2020"
  )
```

```{r}
#| label: factcheck1-viz1
#| code-fold: true
#| fig-width: 12
#| fig-height: 6

# highlight the 160M threshold
ces_data |> 
  mutate(year = year(date)) |>  # create year column first
  filter(year >= 2015) |>
  ggplot(aes(x = date, y = level / 1000)) +
  geom_line(linewidth = 1, color = "steelblue") +
  geom_hline(yintercept = 160, linetype = "dashed", color = "red", linewidth = 1) +
  geom_vline(xintercept = as.Date("2020-01-01"), linetype = "dotted", color = "darkgreen", linewidth = 1) +
  annotate("text", x = as.Date("2020-01-01"), y = 145, 
           label = "Trump's claim\n(Jan 2020)", hjust = 1.1, size = 3.5, color = "darkgreen") +
  annotate("text", x = as.Date("2018-06-01"), y = 160.5, 
           label = "160 Million threshold", hjust = 0, size = 3.5, color = "red") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n%Y") +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  labs(
    title = "US Employment 2015-2020: Was 160 Million Unprecedented?",
    subtitle = "red dashed line shows 160M threshold; green line marks Trump's claim date",
    x = "",
    y = "employment (millions)",
    caption = "Employment exceeded 160M multiple times before January 2020"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```


```{r}
#| label: factcheck1-viz2
#| code-fold: true
#| fig-width: 12
#| fig-height: 6

# compare peak employment by president
presidential_peaks <- ces_with_president |>
  group_by(president) |>
  summarise(
    peak_employment = max(level, na.rm = TRUE),
    peak_date = date[which.max(level)],
    .groups = "drop"
  ) |>
  filter(!is.na(president))

ggplot(presidential_peaks, aes(x = reorder(president, peak_employment), 
                                y = peak_employment / 1000)) +
  geom_col(aes(fill = president == "Trump I"), alpha = 0.8) +
  geom_text(aes(label = format(round(peak_employment / 1000, 1), nsmall = 1)),
            hjust = -0.2, size = 3.5) +
  scale_fill_manual(values = c("FALSE" = "gray70", "TRUE" = "coral"), guide = "none") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Peak Employment by President",
    subtitle = "Trump I highlighted; employment in millions",
    x = "",
    y = "peak employment (millions)",
    caption = "Trump's peak employment was not unprecedented in absolute terms"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```


#### Fact Check #2: Trump's Rigged Revision Claim
"Bureau of Labor Statistics employment numbers "were rigged" when the agency revised them "down by almost 900,000 jobs" after the 2024 election." - Donald Trump, August 5, 2025


```{r}
#| label: factcheck2-stats
#| code-fold: true
#| code-summary: "show analysis of revision rigging claim"

# analyze 2024 revisions
revisions_2024 <- ces_combined |>
  filter(year == 2024)

# calculate total revision for 2024
total_revision_2024 <- sum(revisions_2024$revision, na.rm = TRUE)

# compare to historical patterns
revision_by_year <- ces_combined |>
  group_by(year) |>
  summarise(
    total_annual_revision = sum(revision, na.rm = TRUE),
    mean_revision = mean(revision, na.rm = TRUE),
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    n_negative = sum(revision < 0, na.rm = TRUE),
    n_positive = sum(revision > 0, na.rm = TRUE),
    .groups = "drop"
  )

# show recent years
revision_by_year |>
  filter(year >= 2020) |>
  mutate(
    Year = year,
    `Total Revision` = format(total_annual_revision, big.mark = ","),
    `Mean Revision` = round(mean_revision, 1),
    `Mean Abs Revision` = round(mean_abs_revision, 1),
    `# Negative` = n_negative,
    `# Positive` = n_positive
  ) |>
  select(Year, `Total Revision`, `Mean Revision`, `Mean Abs Revision`, 
         `# Negative`, `# Positive`) |>
  datatable(
    rownames = FALSE,
    options = list(dom = 't', ordering = FALSE),
    caption = "Annual Revision Patterns 2020-2025"
  )
```

```{r}
#| label: factcheck2-test
#| code-fold: true

# test if 2024 revisions are significantly different from pre-2024 average
test_2024_unusual <- ces_combined |>
  mutate(is_2024 = year == 2024) |>
  t_test(abs_revision ~ is_2024,
         order = c("TRUE", "FALSE"),
         alternative = "two.sided")

test_2024_unusual |>
  mutate(
    Test = "2024 Revisions vs Historical Average",
    `T-Statistic` = round(statistic, 3),
    `P-Value` = format.pval(p_value, digits = 3),
    `Mean Difference` = round(estimate, 2),
    Conclusion = ifelse(p_value < 0.05,
                       "2024 revisions ARE statistically unusual",
                       "2024 revisions are NOT statistically unusual")
  ) |>
  select(Test, `T-Statistic`, `P-Value`, `Mean Difference`, Conclusion) |>
  datatable(
    rownames = FALSE,
    options = list(dom = 't'),
    caption = "Statistical Test: Are 2024 Revisions Unusual?"
  )
```
This is also very easily to fact check and disprove. Total 2024 revisions were approximately **`r format(total_revision_2024, big.mark = ",")`** thousand, not "almost 900,000"

"Rigged" implies intentional bias but statistical testing shows 2024 revisions are **not** unusual given that our p-value is **`r format(total_revision_2024, digits = 3)`**

It's important to note that revisions of similar magnitude occurred under both Democratic and Republican administrations, with no systematic partisan pattern. BLS revisions follow a regular schedule of being released every month, independent of elections. The revision process is transparent, methodologically sound, and overseen by career statisticians, making it apolitical.

The claim that revisions were "rigged" is completely false with no statistical or procedural evidence to support it.
Our analysis concludes that this claim is pants on fire, in line with PolitiFact.

```{r}
#| label: factcheck2-viz1
#| code-fold: true
#| fig-width: 12
#| fig-height: 6

# plot total annual revisions over time
ggplot(revision_by_year, aes(x = year, y = total_annual_revision)) +
  geom_col(aes(fill = year == 2024), alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -900, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 1985, y = -900, 
           label = "Trump's claimed -900k", vjust = -0.5, size = 3.5, color = "red") +
  scale_fill_manual(values = c("FALSE" = "gray70", "TRUE" = "coral"), guide = "none") +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(1980, 2024, 5)) +
  theme_minimal() +
  labs(
    title = "Total Annual CES Revisions 1979-2024",
    subtitle = "2024 highlighted; positive values = upward revisions, negative = downward",
    x = "year",
    y = "total annual revision (thousands)",
    caption = "2024 revisions are within normal historical range"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```


```{r}
#| label: factcheck2-viz2
#| code-fold: true
#| fig-width: 12
#| fig-height: 6

# analyze revisions by party
party_revisions <- ces_with_president |>
  filter(!is.na(party)) |>
  group_by(party, president) |>
  summarise(
    mean_revision = mean(revision, na.rm = TRUE),
    mean_abs_revision = mean(abs_revision, na.rm = TRUE),
    total_revision = sum(revision, na.rm = TRUE),
    n_months = n(),
    .groups = "drop"
  )

ggplot(party_revisions, aes(x = president, y = mean_revision, fill = party)) +
  geom_col(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(
    values = c("D" = "blue", "R" = "red"),
    labels = c("D" = "Democratic", "R" = "Republican"),
    name = "Party"
  ) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Average Monthly Revision by President",
    subtitle = "positive values = upward revisions; negative = downward revisions",
    x = "",
    y = "average monthly revision (thousands)",
    caption = "No systematic bias favoring or disfavoring either party"
  ) +
  theme(plot.title = element_text(face = "bold", size = 14))
```


